---
title: "ADHD and MCI cognitive functions in R"
output: pdf_document
---

Assessing cognitive and behavioral differences between older subjects with ADHD and MCI in R.

## Required packages

```{r}
require(readxl)
require(dplyr)
require(ggplot2)
require(viridis)
require(reshape2)
require(GGally)
require(cowplot)
require(tidyr)
```

## Data preprocessing

```{r}
clean_df <- read_excel(path = "data/ADHD_MC.xlsx")
clean_df[clean_df == 999] <- NA
clean_df <- clean_df %>%
  mutate_all(~ifelse(is.na(.), mean(., na.rm = TRUE), .))
clean_df
```

```{r}
#Missing values from the adhd_df
nrow(clean_df)
sum(is.na(clean_df))
last_cols <- clean_df[c('cubost','TMTAz','TMTBz')]
colSums(is.na(last_cols))
```

```{r}
str(clean_df)
```

There is one empty column in this dataset.

## Data visualisation

```{r}
pastel_colors <- c("#A6CEE3", "#B2DF8A")
labels <- c("Women", "Men")
ggplot(clean_df, aes(x = factor(SEXO), fill = factor(SEXO))) +
  geom_bar() +
  labs(title = "Ratio between men and women", x = "Sex", y = "Number") +
  scale_x_discrete(labels = c("Women", "Men")) +
  scale_fill_manual(values = pastel_colors, labels = labels) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r}
sex <- table(clean_df$SEXO)
women <- sex[1] / (sex[1] + sex[2]) * 100
men <- 100 - women
message <- sprintf("There were %.2f%% females.", women)
cat(message, "\n")
```

```{r}
pastel_colors <- c("#A6CEE3", "#B2DF8A", "#D8BFD8")
labels <- c("control", "MCI", "ADHD")

ggplot(clean_df, aes(x = factor(DX), fill = factor(DX))) +
  geom_bar() +
  labs(title = "Ratio between clinical groups", x = "Clinical group", y = "Number") +
  scale_x_discrete(labels = c("control", "MCI", "ADHD")) +
  scale_fill_manual(values = pastel_colors, labels = labels) +  
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  
    panel.grid.major = element_blank(),      
    panel.grid.minor = element_blank()       
  )
```

```{r}
groups <- table(clean_df$DX)
control <- groups[1]/(groups[1]+groups[2]+groups[3])*100
mci <- groups[2]/(groups[1]+groups[2]+groups[3])*100
adhd <- 100 - (control+mci)

variables <- c("control", "mci", "adhd")
labels <- c("controls", "MSI patients", "ADHD patients")

for (i in seq_along(variables)) {
  message <- sprintf("There were %.2f%% %s.", get(variables[i]), labels[i])
  cat(message, "\n")
}
```

```{r}
clean_df %>%
  mutate (Anxiety = factor(Anxiety,
                      levels = c('0','1'),
                      labels = c('anxiety', 'no anxiety'))) %>%
  mutate (DX = factor(DX,
                      levels = c('0','1','2'),
                      labels = c('control', 'MSI','ADHD'))) %>%
  mutate (SEXO = factor(SEXO,
                      levels = c('0','1'),
                      labels = c('women', 'men'))) %>%
  ggplot(aes(IDADE, `MEEM total`))+
  labs(x = "Mental age", y = "Total MEEM")+
  geom_point(aes(color=DX), alpha = 0.5)+
  geom_smooth(method = 'lm')+
  facet_grid(SEXO~Anxiety)
```

```{r}
ggplot(clean_df, aes(x = factor(DX), y = Depression, fill = DX)) +
  geom_boxplot(alpha=0.5) +
  labs(x = "Clinical group", y = "Depression")+ 
  geom_point(aes(color=DX), alpha = 0.5)+
  geom_smooth()
```

```{r}
summary_depression <- aggregate(Depression ~ DX, data = clean_df, function(x) c(mean = mean(x), sd = sd(x)))
summary_depression2 <- do.call(data.frame, summary_depression)
summary_depression2
```

```{r}
ggplot(summary_depression2, aes(x = DX, y = Depression.mean, fill = DX)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = Depression.mean - Depression.sd, ymax = Depression.mean + Depression.sd), width = 0.2, position = position_dodge(width = 0.8)) +
  labs(title = "Depression Summary by Clinical Group", x = "Clinical group", y = "Mean Depression Score") +
  theme_minimal() +
  geom_point(aes(color=DX), alpha = 0.5)+
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none" 
  )
```

There is higher depression rate in ADHD patients, but the standard deviation values show that there is no statistical significance.

```{r}
clean_df %>%
  filter(cubos != '999') %>%                         #remove the outlier
  ggplot(aes(x = factor(DX), y = cubos, fill = DX)) +
  geom_boxplot(alpha=0.5) +
  labs(x = "Clinical group", y = "Cubos")+ 
  geom_point(aes(color=DX), alpha = 0.5)+
  geom_smooth()
```

```{r}
summary_anxiety <- aggregate(Anxiety ~ DX, data = clean_df, function(x) c(mean = mean(x), sd = sd(x)))
summary_anxiety2 <- do.call(data.frame, summary_anxiety)
summary_anxiety2
```

```{r}
ggplot(summary_anxiety2, aes(x = DX, y = Anxiety.mean, fill = DX)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = Anxiety.mean - Anxiety.sd, ymax = Anxiety.mean + Anxiety.sd), width = 0.2, position = position_dodge(width = 0.8)) +
  labs(title = "Anxiety Summary by Clinical Group", x = "Clinical group", y = "Mean Anxiety Score") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none" 
  )
```

There is higher anxiety rate in ADHD patients, but the standard deviation values show that there is no statistical significance.

```{r}
ggplot(clean_df, aes(x = factor(DX), y = a1a5, fill = DX)) +
  geom_boxplot(alpha=0.5) +
  labs(x = "Clinical group", y = "a1a5")+ 
  geom_point(aes(color=DX), alpha = 0.5)+
  geom_smooth()
```

```{r}
data<-dplyr::select(clean_df, DX, Animais, FAS, TMTA, A7)
print(data)
```

```{r}
data %>%
  ggpairs(aes(color = as.factor(DX), alpha = 0.7),
    upper = list(continuous = wrap("points", size = 2.5))) +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )
```

```{r}
hist.IDADE <- ggplot(data = clean_df, aes(IDADE)) +
  geom_histogram()

hist.GAI <- ggplot(data = clean_df, aes(GAI)) +
  geom_histogram()

hist.GDS <- ggplot(data = clean_df, aes(GDS)) +
  geom_histogram()

hist.ESC <- ggplot(data = clean_df, aes(ESC)) +
  geom_histogram()

labels = c('age', 'gai','gds','ESC')

plot_grid(hist.IDADE, hist.GAI, hist.GDS, hist.ESC, align = "hv")
```

## Statistically significant data

```{r}
significant_columns <- c()

for (column in 3:32) {
  model <- aov(DX ~ factor(clean_df[[column]]), data = clean_df)
  summary_result <- summary(model)
  if (summary_result[[1]]$`Pr(>F)`[1] < 0.05) {
    cat("Column:", colnames(clean_df)[column], "has a significant difference.\n")
    significant_columns <- c(significant_columns, colnames(clean_df)[column])
  }
}
```

```{r}
# Tworzenie ramki danych z danymi do wykresu
mean_data <- clean_df %>%
  group_by(DX) %>%
  summarise(across(all_of(significant_columns), mean))

# Przygotowanie danych do wykresu w formacie "dłuższym"
mean_data_long <- pivot_longer(mean_data, cols = -DX, names_to = "Column", values_to = "Mean")

# Tworzenie wykresu kolumnowego
ggplot(mean_data_long, aes(x = Column, y = Mean, fill = factor(DX))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Means for Significant Columns with respect to DX", y = "Mean", fill = "DX") +
  theme_minimal()
```

```{r}
significant_data<-dplyr::select(clean_df, DX, significant_columns)
print(significant_data)
```

## Linear model

```{r}
df.lm1 <- lm(DX ~ ., data = significant_data)
summary(df.lm1)
```

```{r}
par(mfrow = c(2, 2))
plot(df.lm1, pch = 20)
```

-   On residuals vs fitted we can see that there are 3 groups, there are no non-linear patterns.
-   The Q-Q plot shows that the data does not deviate from normal.
-   On scale-location points are more or less near a red line - data is homosedantic.
-   On the last plot we can see that there is one outlier - 107.
